---
title: "EDA"
author: "Kyle Hawley"
date: "2023-05-03"
output: pdf_document
---

TIME
US STATES

MONEY
ACCEPTANCE
INDUSTRY


```{r}
df <- read.csv("./data_csv/processed_h-1b_data.csv")
```

Useful function for changing NAICS_prefixes to labels

```{r}
naics_conversion <-  array()
naics_conversion[11] <- "Agriculture, Forestry, etc."
naics_conversion[21] <- "Mining"
naics_conversion[22] <- "Utilities"
naics_conversion[23] <- "Construction"
naics_conversion[31] <- "Manufacturing"
	
naics_conversion[42] <- "Wholesale Trade"
naics_conversion[44] <- "Retail Trade"
naics_conversion[48] <- "Transportation"
naics_conversion[51] <- "Information"
naics_conversion[52] <- "Finance and Insurance"

naics_conversion[53] <- "Real Estate"
naics_conversion[54] <- "Scientific and Technical Services"
naics_conversion[55] <- "Management of Companies"
naics_conversion[56] <- "Administrative"
naics_conversion[61] <- "Educational Services"
	
naics_conversion[62] <- "Health Care"
naics_conversion[71] <- "Arts"
naics_conversion[72] <- "Food Services"
naics_conversion[81] <- "Other Services"
naics_conversion[92] <- "Public Administration"

code_to_label <- function(code) {
	return(naics_conversion[as.numeric(code)])
}

```




Average Salary

```{r}
library(data.table)

# Data table the dataframe
dt <- data.table(df)

# Create a sequence of dates to include in the output
# months_seq <- seq(min(dt$months_num, na.rm=T), max(dt$months_num, na.rm=T), by = 1)
quarters_seq <- seq(min(dt$quarters_num, na.rm=T), max(dt$quarters_num, na.rm=T), by = 1)

# Use CJ() to create new data tables with all the months and quarters
# dt_complete_mo <- CJ(months_num = months_seq)
dt_complete_qt <- CJ(quarters_num = quarters_seq)

# Join the complete data table with the original data table to fill in missing values with NA
# dt_joined_mo <- merge(dt_complete_mo, dt, by = "months_num", all.x = TRUE, roll = TRUE)
dt_joined_qt <- merge(dt_complete_qt, dt, by = "quarters_num", all.x = TRUE, roll = TRUE)

# Calculate the percentage of accepted values for each date and store the result in a variable
# average_salary_mo <- dt_joined_mo[, mean(wage_scaled, na.rm = TRUE), by = months_num][, V1]
average_salary_qt <- dt_joined_qt[, mean(wage_scaled, na.rm = TRUE), by = quarters_num][, V1]

# average_salary_mo_norm <- average_salary_mo - mean(average_salary_mo, na.rm=T)
average_salary_qt_norm <- average_salary_qt - mean(average_salary_qt, na.rm=T)

# length(months_seq)
# length(average_salary_mo)

# length(quarters_seq)
# length(average_salary_qt)

# plot(months_seq, average_salary_mo, type="l")
plot(quarters_seq, average_salary_qt, type="l", main="Overall Mean Wage", 
		 xlab="Quarters since 2009", ylab="Wage")
plot(quarters_seq, average_salary_qt_norm, type='l', main="Mean Wage Normalized", ylab="", xlab="Quarters since 2009")
```

Certification Rates

```{r}
for (code in sort(unique(df$naics_prefix))) {
	dt_code <- dt[naics_prefix == code]
	
	percent <- sum(dt_code$is_accepted, na.rm=T) / nrow(dt_code)
	
	print(paste0(code_to_label(code), ": ", round(percent,3)))
}
```

# Average salary by Code

```{r}

for (code in sort(unique(df$naics_prefix))) {
	dt_code <- dt[naics_prefix == code]
	dt_joined <- merge(dt_complete_qt, dt_code, by = "quarters_num", all.x = TRUE, roll = TRUE)
	
  ind_wage_mean <- dt_joined[, mean(wage_scaled, na.rm = TRUE), by = quarters_num][, V1]
  
  plot(quarters_seq, ind_wage_mean, main=code_to_label(code), type="l", xlab = "Quarters since 2009", ylab="Mean Wage ($)", ylim=c(40000, 110000))
  points(quarters_seq, average_salary_qt, col="red", type="l")
  legend("bottomright", legend = c("overall mean", "industry mean"), col = c("red", "black"), lwd = 2)
}

```

Average wage grouped by state

```{r}
for (state_obs in unique(df$state)) {
	
	dt_w_state <- dt[state == state_obs]
	dt_joined <- merge(dt_complete_qt, dt_w_state, by = "quarters_num", all.x = TRUE, roll = TRUE)
	
  state_wage_mean <- dt_joined[, mean(wage_scaled, na.rm = TRUE), by = quarters_num][, V1]
  
  plot(quarters_seq, state_wage_mean, main=state_obs, type="l",
  		 xlab="Quarters since 2009", ylab="Mean Wage ($)")
  points(quarters_seq, average_salary_qt, col="red", type="l")
  legend("bottomright", legend = c("overall mean", "state mean"), col = c("red", "black"), lwd = 2)
}
```


Number of cases recieved vs. average salary

```{r}
# Join the complete data table with the original data table to fill in missing values with NA
dt_joined <- merge(dt_complete_qt, dt, by = "quarters_num", all.x = TRUE, roll = TRUE)

# Calculate the percentage of accepted values for each date and store the result in a variable
count_by_qt <- dt_joined[, length(naics_prefix), by = quarters_num][, V1]

plot(quarters_seq, count_by_qt, type="l", xaxt="n")
axis(side = 1, at = seq(1, 180, 4), labels = seq(1, 180, 4), tick = TRUE)

points(quarters_seq, average_salary_qt, type="l", col="red")
```


#Maps
heatmap of applications by state by year

Getting maps up and off the ground

```{r}
library(usmap)
library(ggplot2)

dt <- data.table(df)
dt_2019 <- dt[year == 2019]

dt_state <- dt_2019[, length(date), by = state][, V1, keyby = state]

popdf <- read.csv("./data_csv/pop.csv")
popdt <- data.table(popdf)
colnames(popdt) = c("state", "year", "pop")
popdt19 <- popdt[year == 2019]

dt_state <- merge(dt_state, popdt19, by = "state")
dt_state[, percap := V1 / pop]
dt_state[, percaplog := log(percap)]

p1 <- plot_usmap(data = dt_state, values = "pop", color = "red") + 
  scale_fill_continuous(name = "Counts", low='white',high='blue') + 
  theme(legend.position = "right")

p2 <- plot_usmap(data = dt_state, values = "percap", color = "red") + 
  scale_fill_continuous(name = "Counts", low='white',high='blue') + 
  theme(legend.position = "right")

p3 <- plot_usmap(data = dt_state, values = "percaplog", color = "red") + 
  scale_fill_continuous(name = "Counts", low='white',high='blue') + 
  theme(legend.position = "right")

plot(p1)
plot(p2)
plot(p3)

```


US Map applications per capita
```{r}
popdf <- read.csv("./data_csv/pop.csv")
popdt <- data.table(popdf)
colnames(popdt) = c("state", "year", "pop")
popdt_year <- popdt[year == 2019]

my_colors <- scale_fill_continuous(name = "Applications per capita log scale", limits = c(-11, -3), low='white', high='blue')


for (year_ind in sort(unique(dt$year))) {
	dt_year <- dt[year == year_ind]

	dt_year <- dt_year[, length(date), by = state][, V1, keyby = state]
	
	dt_year <- merge(dt_year, popdt_year, by = "state")
	dt_year[, percap := V1 / pop]
	dt_year[, percaplog := log(percap)]
		
	p1 <- plot_usmap(data = dt_year, values = "percaplog", color = "red") + 
  	my_colors + 
  	theme(legend.position = "right") +
		ggtitle(year_ind)
	
	plot(p1)
}

```


Not per cap Maps
```{r}
popdf <- read.csv("./data_csv/pop.csv")
popdt <- data.table(popdf)
colnames(popdt) = c("state", "year", "pop")
popdt_year <- popdt[year == 2019]

my_colors <- scale_fill_continuous(name = "Applications per capita log scale", low='white', high='blue')


for (year_ind in sort(unique(dt$year))) {
	dt_year <- dt[year == year_ind]

	dt_year <- dt_year[, length(date), by = state][, V1, keyby = state]
	
	dt_year <- merge(dt_year, popdt_year, by = "state")
	dt_year[, percap := V1 / pop]
	dt_year[, percaplog := log(percap)]
		
	p1 <- plot_usmap(data = dt_year, values = "V1", color = "red") + 
  	my_colors + 
  	theme(legend.position = "right") +
		ggtitle(year_ind)
	
	plot(p1)
}

```

USmap wages 

```{r}
# pdf("./pdfs/meanWage_by_state_map.pdf", width=12, height = 10)
my_colors <- scale_fill_continuous(name = "Mean Wage", limits=c(30000, 140000), low='white', high='blue')

for (year_ind in sort(unique(dt$year))) {
	
	dt_year <- dt[year == year_ind]
	dt_year <- dt_year[, mean(wage, na.rm=T), by = state][, V1, keyby = state]
	dt_year[, logwage := log(V1)]
	
	p1 <- plot_usmap(data = dt_year, values = "V1", color = "red") + 
  	# scale_fill_continuous(name = "wage", low='white',high='blue') +
		my_colors + 
  	theme(legend.position = "right") +
		ggtitle(year_ind)
	
	plot(p1, main=year_ind)
	
}
# dev.off()
```


#Back to wages
Normalized ish

```{r}
for (code in sort(unique(df$naics_prefix))) {
	
	dt_w_code <- dt[naics_prefix == code]
	dt_joined <- merge(dt_complete_qt, dt_w_code, by = "quarters_num", all.x = TRUE, roll = TRUE)
	
  ind_wage_mean <- dt_joined[, mean(wage_scaled, na.rm = TRUE), by = quarters_num][, V1]
  ind_wage_mean_norm <- ind_wage_mean - mean(ind_wage_mean, na.rm=T)
  
  plot(quarters_seq, ind_wage_mean_norm, main=code_to_label(code), type="l", xlab = "Quarters since 2009", ylab="Mean Wage ($)")
  points(quarters_seq, average_salary_qt_norm, col="red", type="l")
  legend("bottomright", legend = c("overall mean", "ind mean"), col = c("red", "black"), lwd = 2)
}
```


```{r}
out <- loess.reg(quarters_seq, average_salary_qt_norm, 1)
plot.loess(quarters_seq, average_salary_qt_norm, out, "LOESS on Mean Wage" )

# for (i in seq(1, 60, by=4)) {
# 	abline(v=i, col='gray')
# }


```


By month

```{r}
out <- loess.reg(months, average_salary_mo_norm, 1)

for (code in sort(unique(df$naics_prefix))) {
	
	dt_w_code <- dt[naics_prefix == code]
	dt_joined <- merge(dt_complete_mo, dt_w_code, by = "months_num", all.x = TRUE, roll = TRUE)
	
  ind_wage_mean <- dt_joined[, mean(wage_scaled, na.rm = TRUE), by = months_num][, V1]
  ind_wage_mean_norm <- ind_wage_mean - mean(ind_wage_mean, na.rm=T)
  
  plot.loess(months, average_salary_mo_norm, out, title=code)
  points(months, ind_wage_mean_norm, main=code, type="l", xlab = "qt since 2008", ylab="mean wage")
  # points(months, average_salary_norm, col="red", type="l")
  legend("bottomright", legend = c("overall mean", "ind mean"), col = c("red", "black"), lwd = 2)
}
```


By quarters

A note that number 54 makes up 62% of the data

```{r}
for (code in sort(unique(df$naics_prefix))) {
	dt_code <- dt[naics_prefix == code]
	dt_joined <- merge(dt_complete_qt, dt_code, by = "quarters_num", all.x = TRUE, roll = TRUE)
	
  ind_wage_mean <- dt_joined[, mean(wage_scaled, na.rm = TRUE), by = quarters_num][, V1]
  ind_wage_mean_norm <- ind_wage_mean - mean(ind_wage_mean, na.rm=T)
  
  # if I were keeping values positive
  # average_salary_qt_norm <- average_salary_qt - (mean(average_salary_qt) - mean(ind_wage_mean, na.rm=T))
  
  out <- loess.reg(quarters_seq, average_salary_qt_norm, 1)
  plot.loess(quarters_seq, average_salary_qt_norm, out, title=code_to_label(code), "Quarters since 2009", "Mean Wage ($)")
  
  points(quarters_seq, ind_wage_mean_norm, type="l", lwd=2)
  
#   Poisson
# 	c_upper <- qpois(0.975, ind_wage_mean)
# 	c_lower <- qpois(0.025, ind_wage_mean)
#   
#   points(quarters, c_upper, col="gray", type="l")
#   points(quarters, c_lower, col="gray", type="l")
  
  # simple sd
  # sd <- sd(ind_wage_mean, na.rm=T)
  # points(quarters, ind_wage_mean + (2*sd), col='gray', type='l')
  # points(quarters, ind_wage_mean - (2*sd), col='gray', type='l')
  
  
  simple_linear_reg <- lm(ind_wage_mean ~ quarters_seq)
  linear_reg_sd <- sd(simple_linear_reg$residuals)
  
  points(quarters_seq, (ind_wage_mean_norm + (2*linear_reg_sd)), col='gray', type='l')
  points(quarters_seq, (ind_wage_mean_norm - (2*linear_reg_sd)), col='gray', type='l')
  
  # for (i in seq(1, 60, by=4)) {
  # 	abline(v=i, col="gray", lwd=1)
  # }
}
```

by state

```{r}
for (state_obs in sort(unique(df$state))) {
	dt_w_state <- dt[state == state_obs]
	dt_joined <- merge(dt_complete_qt, dt_w_state, by = "quarters_num", all.x = TRUE, roll = TRUE)
	
  state_wage_mean <- dt_joined[, mean(wage_scaled, na.rm = TRUE), by = quarters_num][, V1]
  state_wage_mean_norm <- state_wage_mean - mean(state_wage_mean, na.rm=T)
  
  plot.loess(quarters, average_salary_qt_norm, out, state_obs)
  points(quarters, state_wage_mean_norm, main=state_obs, type="l", lwd=2)
}
```












